Gravitational wave sources in our Galactic backyard: LISA predictions for BH and NS binaries

### 2021-06-14 paper/tex/sections/discussion.tex
finish/improve pulsar subsection

\subsection{Matching LISA detection to radio pulsars}\label{sec:pulsar_matching}

>>>>

\subsection{Matching LISA detections to pulsars with SKA}\label{sec:pulsar_matching}
Since the vast majority of the LISA detectable population of DCOs will not merge for many years, the main form of electromagnetic counterpart for the this population is pulsars. Therefore, for this section we focus only on BHNSs and NSNSs since no BHBH system will contain a pulsar. The joint detection of a binary pulsar with LISA and SKA would not only help to constrain the parameters of the binary, but also enable investigation of other compact object physics. A pulsar(PSR)+BH can be provide stringent tests of theories of gravity, in particular the ``No-hair theorem'' \citep{Keane+2015}. Altneratively, an ultrarelativstic PSR+NS system could be used to measure the neutron star equation of state up to an order of magniute better than other proposed observations \citep{Kyutoku+2019, Thrane+2020}.

In this section, we perform some back-of-the-envelope calculations to order to estimate how we can use SKA to match pulsars to LISA detections.

First, we can consider how many pulsars SKA is likely to detect. \citet{Keane+2015} uses PSRPOPPy \citep{Bates+2014} to simulate the Milky Way pulsar population. They find that for SKA-1, approximately $10000$ pulsars will be discovered. The second phase of SKA, which should also have been approved and be in operation by the time of the LISA mission, would yield a total of $35000$-$41000$ pulsars \citep{Keane+2015}, where we use the average, $38000$, in further estimates below. Moreover, we are only interested in pulsars that are part of a binary system. We can estimate this pulsar binary fraction as the fraction of known pulsars that are in binaries using the ATNF Pulsar Catalogue\footnote{\url{https://www.atnf.csiro.au/research/pulsar/psrcat}} \citep{Manchester+2005}. $290$ of the $2872$ currently known pulsars are in binary systems and thus we can estimate the binary fraction of pulsars as $10\%$.

Next, we can find the total number of pulsars SKA will detect in a patch on the sky. The total sky area that the SKA mission covers is approximately $7200 \unit{deg^2}$, which is calculated by integrating over the sky for all Galactic longitudes and Galactic latitudes limited to $\abs{b} < 10^\circ$, which are the limits on SKA-mid \citep{Keane+2015}. If we assume that the pulsars are found uniformly across the sky, this means that roughly $0.14$ and $0.49$-$0.57$ binary pulsars are expected per square degree for SKA-1 and SKA-2 respectively. Note that the assumption assumption of a uniform distribution is not realistic as pulsars will tend to be far more concentrated in the Galactic centre but we use it to provide an upper bound on these estimates.

Alternatively, we could write that we expect a single pulsar per $7.2 \unit{deg^2}$ and $2.1$-$1.8 \unit{deg^2}$ for SKA-1 and SKA-2 respectively, which corresponds to angular resolutions of $\sigma_\theta = 1.51^\circ$ and $\sigma_\theta = 0.82$-$0.76^\circ$. Given these estimates, and by considering the last panel of Fig.~\ref{fig:fiducial_pdf_distributions}, approximately $10$ and $6$ (for SKA-1 and SKA-2) DCOs containing NSs will be localised well enough such that, \textit{if} the NS is a pulsar, SKA can unambiguously match it to the radio signal.

If there is more than one pulsar in the region given by the LISA sky localisation, one can compare the measured parameters of the system in LISA and SKA. Both SKA and LISA will measure the orbital frequency to high precision, as well as the time derivative of the frequency and chirp mass to a lesser precision, of each of these systems. Therefore, one could perform a targetted search with SKA that checks the sky location given by LISA and only looking for binary pulsars with orbital frequencies within the errors. If there was \textit{still} more than one possible pulsar one could then also check against the chirp mass. In this way, it would still be possible to get a joint detection between SKA and LISA even when the sky area implied by the LISA detection contains more than one pulsar.

In order to assess the efficacy of this method, we would need to know the probability that two random binary pulsars would have orbital frequencies and chirp masses close enough that one could not tell which pulsar matches the LISA detection. This would require simulating the SKA population of pulsars with a code such as PSRPOPPy to find the frequency and chirp mass distribution and thus is beyond the scope of this paper. However, the uncertainty on the orbital frequency of a binary on the detection threshold ($\rho = 7$) for a 4-year LISA mission is $2.5 \times 10^{-9} \unit{Hz}$ and $1.0 \times 10^{-9} \unit{Hz}$ for a 10-year mission (calculated using Eq.~\ref{eq:f_orb_unc}). Therefore, we expect that SKA could likely isolate the correct binary pulsar to match to a LISA detection even when several are present in the sky localisation region.

### 2021-06-14 paper/tex/sections/results.tex
finish/improve pulsar subsection

We calculate the frequency uncertainties using \citet{Takahashi+2002}, such that
\begin{align}
    \frac{\Delta f_{\rm orb}}{f_{\rm orb}} &= 4 \sqrt{3} \cdot \frac{1}{\rho} \frac{1}{T_{\rm obs}} \frac{1}{f_{\rm orb}}, \\
    \frac{\Delta \dot{f}_{\rm orb}}{\dot{f}_{\rm orb}} &= 6 \sqrt{5} \cdot \frac{1}{\rho} \left(\frac{1}{T_{\rm obs}} \right)^2 \frac{1}{\dot{f}_{\rm orb}},
\end{align}
where $\rho$ is the signal-to-noise ratio and $T_{\rm obs}$ is the LISA mission length. Then we calculate the eccentricity certainty as discussed in Sec.~\ref{sec:ecc_unc} and propagate it so that
\begin{equation}
    \frac{\Delta F(e)}{F(e)} = \Delta e \cdot \frac{(1256 + 1608 e^2 + 111 e^4) e}{96 + 196 e^2 - 255 e^4 - 37 e^6}.
\end{equation}

@@

We plot the angular resolution in the last panel of Fig.~\ref{fig:fiducial_pdf_distributions} as a CDF normalised to the expected detection rate of each DCO. From this plot, we see that many sources can only be resolved to a patch with an area in excess of 100 square degrees \todo{make this positive}. The size of a pencil beam for a 15m diameter SKA dish observing at 1.4 GHz is roughly 0.67 square degrees, corresponding to an angular resolution of $\sigma_\theta = \sqrt{(0.67 / \pi)} = 0.46^\circ$ (approximately the angular size of the moon). Applying this to the plot shows that about 20\% of DCOs can be covered by a single pointing of SKA. We will discuss the prospects of matching LISA detections to radio pulsars with SKA further in Sec.~\ref{sec:pulsar_matching}.

>>>>

We calculate the frequency uncertainties using \citet{Takahashi+2002}, such that
\begin{align}\label{eq:f_orb_unc}
    \frac{\Delta f_{\rm orb}}{f_{\rm orb}} &= 4 \sqrt{3} \cdot \frac{1}{\rho} \frac{1}{T_{\rm obs}} \frac{1}{f_{\rm orb}}, \\
    \frac{\Delta \dot{f}_{\rm orb}}{\dot{f}_{\rm orb}} &= 6 \sqrt{5} \cdot \frac{1}{\rho} \left(\frac{1}{T_{\rm obs}} \right)^2 \frac{1}{\dot{f}_{\rm orb}},
\end{align}
where $\rho$ is the signal-to-noise ratio and $T_{\rm obs}$ is the LISA mission length. Then we calculate the eccentricity certainty as discussed in Sec.~\ref{sec:ecc_unc} and propagate it so that
\begin{equation}
    \frac{\Delta F(e)}{F(e)} = \Delta e \cdot \frac{(1256 + 1608 e^2 + 111 e^4) e}{96 + 196 e^2 - 255 e^4 - 37 e^6}.
\end{equation}

@@

We plot the angular resolution in the last panel of Fig.~\ref{fig:fiducial_pdf_distributions} as a CDF normalised to the expected detection rate of each DCO. From this plot, we see that the majority of sources can be resolved to an angular resolution of 10 square degrees. The size of a pencil beam for a 15m diameter SKA dish observing at 1.4 GHz is roughly 0.67 square degrees, corresponding to an angular resolution of $\sigma_\theta = \sqrt{(0.67 / \pi)} = 0.46^\circ$ (which, for intuition, is approximately the angular size of the moon). Applying this to the plot shows that about 20\% of DCOs can be covered by a single pointing of SKA. We will discuss the prospects of matching LISA detections to radio pulsars with SKA further in Sec.~\ref{sec:pulsar_matching}.
